---
title: "tidy model"
author: "Jack Torres Zuniga"
output: html_document
date: "2025-02-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(modeldata)
```

## Introduction

This R Markdown document demonstrates my ability to use models for data analysis using a dataset collected on crickets.

```{r load data, include=FALSE}
data(crickets, package = "modeldata")
names(crickets)
```

## What is that sound?

In this report, we examine what factors predict a cricket's chirp rate.

```{r summary, echo=FALSE}
summary(crickets)

ggplot(crickets, aes(x = rate)) +
  geom_histogram(bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```

## Temperature affects chirp rate

```{r temp, echo=FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

temp_lm <- lm(rate ~ temp, crickets)
summary(temp_lm)
```

Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases by one degree, the chirp rate increases by approximately 4.2 chirps per minute.

## Species-specific effects of temperature on chirping

```{r species, echo=FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

species_lm <- lm(rate ~ temp + species, crickets)
summary(species_lm)
```
